Appendix to "Integrated nested Laplace approximations for extended latent Gaussian models with application to the Naomi HIV model

### 2022-12-21 src/docs_paper/appendix.Rmd
Complete try at scrubing out T2 and T3 in appendix, and set stage for writing model summary in main text

## Indexing

Let $x$ be a district, located within the Spectrum region $R_x$, $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
Consider the time period $t = T_1$ of the most recent national household survey with HIV testing specified to the nearest calendar quarter.
We index population size $N_{x, s, a, t}$, HIV prevalence $\rho_{x, s, a, t}$, ART coverage $\alpha_{x, s, a, t}$ and annual HIV incidence rate $\lambda_{x, s, a, t}$ by district, sex, age-band and time period.

## Structured random effects

We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) with marginal standard deviation $\sigma > 0$, and $u \sim \text{BYM}2(\sigma, \phi)$ to the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] with marginal standard deviation $\sigma > 0$ comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$.
For both the ICAR and BYM2 models, we follow recommendations [@freni2018note] on scaling, disconnected adjacency graph components, and islands.

@@

We model HIV prevalence by
$$
\text{logit}(\rho_{x, s, a, T_1}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
$$
where $\beta^\rho_0$ is the intercept, $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}}$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are a fixed offset specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
We use the prior distributions $\mathcal{N}(0, 5)$ for all fixed effects ($\beta^\rho_0, \beta_{s}^{\rho, s = \text{M}}$), $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.

@@

We model ART coverage by
$$
\text{logit}(\alpha_{x, s, a, T_1}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + u^\alpha_a + u_a^{\alpha, s = \text{M}} + u^\alpha_x + u_x^{\alpha, s = \text{M}} + u_x^{\alpha, a < 15} + \eta^\alpha_{R_x, s, a} 
$$
with analogous terms to the HIV prevalence model.

@@

We model HIV incidence rate by
$$
\log(\lambda_{x, s, a, t}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x, t}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x, t}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a, t}^\lambda 
$$
where $\beta^\lambda_0$ is the intercept, which is proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}}$ is the log incidence rate ratio for men compared to women, $\rho_{x, t}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by
$$
\rho_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t}},
$$
$\alpha_{x, t}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by
$$
\alpha_{x, t}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{49} N_{x, s, a, t} \cdot \rho_{x, s, a, t}},
$$
$\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ from the EPP model, $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a, t}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output

@@

For women 15-49 the predicted number of ANC clients $\Psi_{x, a, t}$ is a log-linear model
$$
\log (\Psi_{x, a, t}) = \log (N_{x, \text{F}, a, t}) + \psi_{R_x, a, t} + \beta^\psi + u_x^\psi
$$
where $N_{x, \text{F}, a, t}$ are the female population sizes, $\psi_{R_x, a, t}$ are ASFR in Spectrum region $R_x$ at time $t$, $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility, $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects, HIV prevalence $\rho_{x, a, t}^\text{ANC}$ and ART coverage $\alpha_{x, a, t}^\text{ANC}$ among pregnant women modelled with logit-linear models
\begin{align}
\text{logit}(\rho_{x, a, t}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a, t}) + \beta^{\rho^{\text{ANC}}} + \beta_{t \in(T_2, T_3\}}^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + u_{x, t \in\{T_2, T_3\}}^{\rho^{\text{ANC}}} + \eta_{R_x, a, t}^{\rho^{\text{ANC}}} \\
\text{logit}(\alpha_{x, a, t}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a, t}) + \beta^{\alpha^{\text{ANC}}} + \beta_{t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}}+u_{x, t \in\{T_2, T_3\}}^{\alpha^{\text{ANC}}} + \eta_{R_x, a, t}^{\alpha^{\text{ANC}}},
\end{align}
  * $\eta_{R_x, a, t}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$
  * $\beta^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the average differences between population and ANC outcomes at $T_1$ after removing the offset
  * $\beta_{t \in(T_2, T_3\}}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the change in average difference from $T_1$ to $T_2$
  * $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ are district random effects
  * $u_{x, t \in\{T_2, T_3\}}^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_{XT}^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ capture the change in district random effects between times $T_1$ and $T_2$

@@

* Number of PLHIV on ART is $A_{x, s, a, t} = N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}$
* Let $\gamma_{x, x', t} \in [0, 1]$ be the probability that a person of ART residing in district $x$ recieves ART in district $x'$ at time $t$
* We assume that $\gamma_{x, x', t} = 0$ for $x \notin \{x, \text{ne}(x)\}$ i.e. that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$ where $\sim$ is an adjacency relation
  * $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x', t} = 1$
* Model $\gamma_{x, x', t}$ with a multinomial logit model where $\tilde \gamma_{x, t}$ is^[Confused by this notation.] the log odds ratio of seeking ART in the district $x'$ neighbouring $x$
\begin{align}
\tilde \gamma_{x, t} &= \tilde \gamma_0 + u_x^{\tilde \gamma} + u_x^{\tilde \gamma, t = T_2} \\ 
\tilde \gamma_0 &= -4 \\
u_x^{\tilde \gamma} &\sim \mathcal{N}(0, \sigma_X^{\tilde \gamma}) \\
u_x^{\tilde \gamma, t \in \{T_2, T_3\}} &\sim \mathcal{N}(0, \sigma_{XT}^{\tilde \gamma}) 
\end{align}
  * $\tilde \gamma_{x, t}$ is an intercept specifying the prior mean for the log odds of seeking ART in each neighbouring district compared to the home district
  * $u_x^{\tilde \gamma}$ are district random effects
  * $u_x^{\tilde \gamma, t \in \{T_2, T_3\}}$ is^[Shouldn't this just be $t = T_2$ here rather than $t \in \{T_2, T_3\}$?] the change in district level log-odds ratios between $T_1$ and $T_2$
  * Note: only one value estimated for each district, implying no preference between neighbouring districts, and no sex-age specific modelling
* Calculate multinomial probabilities via softmax (n.b. this is the sigmoid function, the inverse of the softmax)
$$
\gamma_{x, x', t} = \frac{1}{1 + \sum_{x^\star \sim x} \exp(\tilde \gamma_{x^\star, t})}
$$
* The number of ART clients who reside in district $x$ and obtain ART in district $x'$ are $A_{x, x', s, a, t} = A_{x, s, a, t} \cdot \gamma_{x, x', t} = N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t} \cdot \gamma_{x, x', t}$
* Total attending ART facilities in district $x'$ is $\tilde A_{x', s, a, t} = \sum_{x \sim x', x = x'} A_{x, x', s, a, t}$

@@

* The proportion of HIV positive adults aware of their HIV status is $v_{x, s, a, t}$
* The proportion of untreated HIV positive adults aware of their HIV status is $\tilde \nu_{x, s, a, t}$ such that
$$
v_{x, s, a, t} = \alpha_{x, s, a, t} + \frac{\tilde v_{R_x, s, a, t}}{1 - \alpha_{x, s, a, t}}
$$

@@

### Notation

* Use $\{x\}$ to denote a collection of districts, $\{s\}$ a collection of sexes and $\{a\}$ a collection of ages for which an outcome is observed

@@

* Let $\nu$ be a household survey occuring at $T_1$
* Furnishes weighted observations for HIV prevalence $\hat \rho_{\{x\}, \{s\}, \{a\}, \nu}$, ART coverage $\hat \alpha_{\{x\}, \{s\}, \{a\}, \nu}$ and proportion recently infected $\hat \kappa_{\{x\}, \{s\}, \{a\}, \nu}$ with respective Kish effective sample sizes $M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu}$ for $\theta \in \{\rho, \alpha, \kappa\}$
* The observed cases are $Y^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} = M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}, \nu}$
* For HIV prevalence we use a binomial working likelihood
\begin{align}
Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\rho_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t}}
\end{align}
* For ART coverage we also use a binomial working likelihood
\begin{align}
Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\alpha_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t} \cdot \alpha_{x, s, a, t}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a, t} \cdot \rho_{x, s, a, t}}
\end{align}
* For recent infections we also use a binomial working likelihood
\begin{align}
Y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{Bin}(M^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu}, \kappa_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\kappa_{x, s, a, t} &= 1 - \exp(- \lambda_{x, s, a, t} \cdot \frac{1 - \rho_{x, s, a, t}}{\rho_{x, s, a, t}} \cdot (\Omega_T - \beta_T ) - \beta_T ) \\
\Omega_T &\sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T}) \\
\beta_T &\sim \mathcal{N}(\beta_{T_0}, \sigma^{\beta_T})
\end{align}
  * $\kappa_{x, s, a, t}$ are the predicted proportion recently infected among HIV positive persons
  * $\Omega_T$ is the mean duration of recent infection (MDRI)
    * Use an informative prior based on the characteristics of the recent infection testing algorithm (RITA)
    * For PHIA surveys^[What about other surveys?] $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$
  * $\beta_T$ is the false recent ratio (FRR)
    * For PHIA surveys $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.^[Is this just assuming that there is no false recency? How is this assumption justified?]

@@

* Include two years of ANC testing data: the year of the most recent survey $Y[T_1]$ and the current year $Y[T_2]$
* $W^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients
* $X^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with ascertained status
* $Y^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients with positive status (either known or tested)
* $Z^\text{ANC}_{\{x\}, Y[t]}$ are the number of ANC clients already on ART prior to first ANC
* Sometimes ANC testing data are only available for part of a given year, for example only the first three quarters. Denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$
* Likelihood for number of ANC clients specified only^[Why?] for year $Y[T_2]$ as
$$
W_{\{x\} Y[T_2]}^{\text{ANC}} \sim \text{Pois} \left( \frac{M_{Y[T 2]}^{\text{ANC}}}{12} \sum_{x \in\{x\}} \sum_{a \in\{15, \ldots 45\}} \Psi_{x, a, T_2} \right)
$$
* Observed number of HIV positive and already on ART among ANC clients at both $Y[T_1]$ and $Y[T_2]$ are modelled by
\begin{align*}
& Y_{\{x\} Y[t]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[t]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 49\}, t}^{\text{ANC}}\right) \\
& Z_{\{x\} Y[t]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[t]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 49\}, t}^{\text{ANC}}\right)
\end{align*}
where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a, t}$
\begin{align*}
\rho_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t}} \\
\alpha_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}} \cdot \alpha_{x, a, t}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a, t} \cdot \rho_{x, a, t}^{\text{ANC}}}
\end{align*}

>>>>

### Indexing

Consider some country, and let $x$ refer to a district located within the Spectrum region $R_x$.
We restrict our attention to the time period $t = T_1$ of the most recent national household survey with HIV testing.
Let $s \in \{\text{F}, \text{M}\}$ be sex, and $a \in \{\text{0-5}, \text{5-10}, \ldots, \text{75-80}, 80+\}$ be five-year age bands.
As short-hand, we write $a = l$ to refer to the age band with lower bound $l$.
We index the following quantities by district, sex, age-band: population size $N_{x, s, a}$, HIV prevalence $\rho_{x, s, a}$, ART coverage $\alpha_{x, s, a}$ and annual HIV incidence rate $\lambda_{x, s, a}$.
Our model operates at this level of granularity, although sometimes the data observed is at an aggregate level.
We use $\{\cdot\}$ to refer to a generic aggregate set over which an observation is made, such as $\{a\} = \{\text{15-19}, \ldots, \text{45-49}\}$ for adults.

### Structured random effects

Consider some generic random effect $u$.
We use $u \sim \text{ICAR}(\sigma)$ to refer to the Besag intrinsic conditional auto-regressive model (ICAR) [@besag1991bayesian] with marginal standard deviation $\sigma > 0$.
For the reparameterised Besag-York-Mollie model (BYM2) [@simpson2017penalising] with marginal standard deviation $\sigma > 0$ we write $u \sim \text{BYM}2(\sigma, \phi)$, where $u$ is comprised of a spatially structured ICAR component with proportion $\phi \in (0, 1)$ and spatially unstructured IID component with proportion $1 - \phi$
We follow recommendations of @freni2018note on scaling, disconnected adjacency graph components, and islands.
The first order auto-regressive model is specified by $u \sim \text{AR}1(\sigma, \phi)$.

@@

We model HIV prevalence by
$$
\text{logit}(\rho_{x, s, a}) = \beta^\rho_0 + \beta_{S}^{\rho, s = \text{M}} + u^\rho_a + u_a^{\rho, s = \text{M}} + u^\rho_x + u_x^{\rho, s = \text{M}} + u_x^{\rho, a < 15} + \eta^\rho_{R_x, s, a} 
$$
where $\beta^\rho_0$ is an intercept term, $\beta_{s}^{\rho, s = \text{M}}$ is the difference in logit prevalence for men compared to women, $u^\rho_a \sim \text{AR}1(\sigma_A^\rho, \phi_A^\rho)$ are age random effects for women, $u_a^{\rho, s = \text{M}}$ are age random effects for the difference in logit prevalence for men compared to women age $a$, $u^\rho_x \sim \text{BYM}2(\sigma_X^\rho, \phi_X^\rho)$ are spatial random effects for women, $u_x^{\rho, s = \text{M}} \sim \text{BYM}2(\sigma_{XS}^\rho, \phi_{XS}^\rho)$ are spatial random effects for the difference in logit prevalence for men compared to women in district $x$, $u_x^{\rho, a < 15} \sim \text{ICAR}(0, \sigma_{XA}^\rho)$ are spatial random effects for the ratio of paediatric prevalence to adult women prevalence, and $\eta^\rho_{R_x, s, a}$ are fixed offsets specifying assumed odds ratios for prevalence outside the age ranges for which data are available.
We use the prior distributions $\mathcal{N}(0, 5)$ for the fixed effects $\beta^\rho_0$ and $\beta_{s}^{\rho, s = \text{M}}$, $\mathcal{N}^{+}(0, 2.5)$ for all standard deviation terms, $\mathcal{U}(-1, 1)$ for all AR1 correlation parameters, and $\text{Beta}(0.5, 0.5)$ for all BYM2 proportion parameters.

@@

We model ART coverage by
$$
\text{logit}(\alpha_{x, s, a}) = \beta^\alpha_0 + \beta_{S}^{\alpha, s = \text{M}} + u^\alpha_a + u_a^{\alpha, s = \text{M}} + u^\alpha_x + u_x^{\alpha, s = \text{M}} + u_x^{\alpha, a < 15} + \eta^\alpha_{R_x, s, a} 
$$
with terms and priors analogous to the HIV prevalence model above.

@@

We model HIV incidence rate by
$$
\log(\lambda_{x, s, a}) = \beta_0^\lambda + \beta_S^{\lambda, s = \text{M}} + \log(\rho_{x}^{\text{15-49}}) + \log(1 - \omega \cdot \alpha_{x}^{\text{15-49}}) + u_x^\lambda + \eta_{R_x, s, a}^\lambda 
$$
where $\beta^\lambda_0$ is an intercept term proportional to the average HIV transmission rate for untreated HIV positive adults, $\beta_S^{\lambda, s = \text{M}}$ is the log incidence rate ratio for men compared to women, $\rho_{x}^{\text{15-49}}$ is the HIV prevalence among adults 15-49 calculated by
$$
\rho_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a}},
$$
$\alpha_{x}^{\text{15-49}}$ is the ART coverage among adults 15-49 calculated by
$$
\alpha_{x}^{\text{15-49}} = \frac{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{s \in \{\text{F}, \text{M}\}} \sum_{a = 15}^{45} N_{x, s, a} \cdot \rho_{x, s, a}},
$$
$\omega$ is the average reduction in HIV transmission rate per 1% increase in population ART coverage and is fixed at $\omega = 0.7$ via the EPP model, $u_x^\lambda \sim \mathcal{N}(0, \sigma^\lambda)$ with $\sigma^\lambda \sim \mathcal{N}^+(0, 1)$ are IID spatial random effects, and $\eta^\lambda_{R_x, s, a}$ specify log incidence rate ratios by sex and age group calculated from Spectrum model output

@@

For women 15-49 the predicted number of ANC clients $\Psi_{x, a}$ is a log-linear model
$$
\log (\Psi_{x, a}) = \log (N_{x, \text{F}, a}) + \psi_{R_x, a} + \beta^\psi + u_x^\psi
$$
where $N_{x, \text{F}, a}$ are the female population sizes, $\psi_{R_x, a}$ are ASFR in Spectrum region $R_x$ at time $t$, $\beta^\psi$ are the log rate ratio for the number of ANC clients relative to the predicted fertility, $u_x^\psi \sim \mathcal{N}(0, \sigma^\psi)$ are district random effects.
HIV prevalence $\rho_{x, a}^\text{ANC}$ and ART coverage $\alpha_{x, a}^\text{ANC}$ among pregnant women modelled with logit-linear models
\begin{align}
\text{logit}(\rho_{x, a}^{\text{ANC}}) &= \text{logit}(\rho_{x, F, a}) + \beta^{\rho^{\text{ANC}}} + u_x^{\rho^{\text{ANC}}} + \eta_{R_x, a}^{\rho^{\text{ANC}}} \\
\text{logit}(\alpha_{x, a}^{\text{ANC}}) &= \text{logit}(\alpha_{x, F, a}) + \beta^{\alpha^{\text{ANC}}} + u_x^{\alpha^{\text{ANC}}} + \eta_{R_x, a}^{\alpha^{\text{ANC}}},
\end{align}

where $\eta_{R_x, a}^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are offsets for the log fertility rate ratios for HIV positive women compared to HIV negative women and for women on ART to HIV positive women not on ART, calculated from Spectrum model outputs for region $R_x$, $\beta^{\theta^{\text{ANC}}}$ for $\theta \in \{\rho, \alpha\}$ are the average differences between population and ANC outcomes after removing the offset, and $u_x^{\theta^{\text{ANC}}} \sim \mathcal{N}(0, \sigma_X^{\theta^{\text{ANC}}})$ for $\theta \in \{\rho, \alpha\}$ are district random effects.

@@

Let $\gamma_{x, x'} \in [0, 1]$ be the probability that a person of ART residing in district $x$ recieves ART in district $x'$.
We assume that $\gamma_{x, x'} = 0$ for $x \notin \{x, \text{ne}(x)\}$ such that individuals seek treatment only in their residing district and its neighbours $\text{ne}(x) = \{x': x' \sim x\}$, where $\sim$ is an adjacency relation and $\sum_{x' \in \{x, \text{ne}(x)\}} \gamma_{x, x'} = 1$.

To model $\gamma_{x, x'}$ for $x \sim x'$ we use a multinomial logistic regression model where $\tilde \gamma_{x, x'}$ is the log odds ratio of seeking ART
\begin{align}
\tilde \gamma_{x, x'} &= \tilde \gamma_0 + u_x^{\tilde \gamma} \\ 
\tilde \gamma_0 &= -4 \\
u_x^{\tilde \gamma} &\sim \mathcal{N}(0, \sigma_X^{\tilde \gamma})
\end{align}
where $\tilde \gamma_0$ is an intercept specifying the prior mean for the log odds of seeking ART in each neighbouring district compared to the home district, and $u_x^{\tilde \gamma}$ are district random effects.
We fix $\tilde \gamma_{x, x} = 0$ and recover the multinomial probabilities using the softmax
$$
\gamma_{x, x'} = \frac{\exp(\tilde \gamma_{x, x'})}{\sum_{x^\star \in \{x, \text{ne}(x)\}} \exp(\tilde \gamma_{x, x^\star})}
$$
Then, given the number of PLHIV on ART $A_{x, s, a} = N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}$, the number of ART clients who reside in district $x$ and obtain ART in district $x'$ are
$$
A_{x, x', s, a} = A_{x, s, a} \cdot \gamma_{x, x'},
$$
and the total attending ART facilities in district $x'$ are
$$
\tilde A_{x', s, a} = \sum_{x \in \{x', \text{ne}(x')\}} A_{x, x', s, a}.
$$

@@

The proportion of HIV positive adults aware of their HIV status is $v_{x, s, a}$, and the proportion of untreated HIV positive adults aware of their HIV status is $\tilde \nu_{x, s, a}$ such that
$$
v_{x, s, a} = \alpha_{x, s, a} + \frac{\tilde v_{R_x, s, a}}{1 - \alpha_{x, s, a}}.
$$

@@

@@

Let $\nu$ be a household survey occurring at $T_1$, furnishing weighted observations $\hat \theta_{\{x\}, \{s\}, \{a\}, \nu}$ for $\theta \in \{\rho, \alpha, \kappa\}$ with respective Kish effective sample sizes $M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu}$, and observed number of cases
$$
Y^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} = M^{\hat \theta}_{\{x\}, \{s\}, \{a\}, \nu} \cdot \hat \theta_{\{x\}, \{s\}, \{a\}, \nu}.
$$

For HIV prevalence and ART coverage we use binomial working likelihoods as follows
\begin{align}
Y^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \rho}_{\{x\}, \{s\}, \{a\}, \nu}, \rho_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\rho_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a}},
\end{align}
and
\begin{align}
Y^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \alpha}_{\{x\}, \{s\}, \{a\}, \nu}, \alpha_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\alpha_{\{x\}, \{s\}, \{a\}, t} &= \frac{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a} \cdot \alpha_{x, s, a}}{\sum_{x \in \{x\}} \sum_{s \in \{s\}} \sum_{a \in \{a\}} N_{x, s, a} \cdot \rho_{x, s, a}}.
\end{align}

For recent infections we also use a binomial working likelihood
\begin{align}
Y^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu} &\sim \text{xBin}(M^{\hat \kappa}_{\{x\}, \{s\}, \{a\}, \nu}, \kappa_{\{x\}, \{s\}, \{a\}, {T_1}}) \\
\kappa_{x, s, a} &= 1 - \exp(- \lambda_{x, s, a} \cdot \frac{1 - \rho_{x, s, a}}{\rho_{x, s, a}} \cdot (\Omega_T - \beta_T ) - \beta_T ) \\
\Omega_T &\sim \mathcal{N}(\Omega_{T_0}, \sigma^{\Omega_T}) \\
\beta_T &\sim \mathcal{N}(\beta_{T_0}, \sigma^{\beta_T})
\end{align}
where $\kappa_{x, s, a}$ are the predicted proportion recently infected among HIV positive persons, $\Omega_T$ is the mean duration of recent infection (MDRI), and $\beta_T$ is the false recent ratio (FRR).
We use an informative prior on $\Omega_T$ based on the characteristics of the recent infection testing algorithm (RITA).
For PHIA surveys this is $\Omega_{T_0} = 130 \text{ days}$ and $\sigma^{\Omega_T} = 6.12 \text{ days}$.
For PHIA surveys we assume there is no false recency, such that $\beta_{T_0} = 0.0$ and $\sigma^{\beta_T} = 0.0$.

@@

We include ANC testing data for the year of the most recent survey $Y[T_1]$.
Let $W^\text{ANC}_{\{x\}, Y[t]}$ be the number of ANC clients, $X^\text{ANC}_{\{x\}, Y[t]}$ the number of those with ascertained status, $Y^\text{ANC}_{\{x\}, Y[t]}$ the number of those with positive status (either known or tested) and $Z^\text{ANC}_{\{x\}, Y[t]}$ the number of ANC clients already on ART prior to first ANC, such that $W^\text{ANC}_{x, Y[t]} \geq X^\text{ANC}_{x, Y[t]} \geq Y^\text{ANC}_{x, Y[t]} \geq Z^\text{ANC}_{x, Y[t]}$
When ANC testing data are only available for part of a given year, we denote $M^\text{ANC}_{Y[t]} \in \{1, \ldots, 12\}$ the number of months of reported data reflected in counts for year $Y[t]$.

<!-- * Likelihood for number of ANC clients specified only^[Why?] for year $Y[T_2]$ as -->
<!-- $$ -->
<!-- W_{\{x\} Y[T_2]}^{\text{ANC}} \sim \text{Pois} \left( \frac{M_{Y[T 2]}^{\text{ANC}}}{12} \sum_{x \in\{x\}} \sum_{a \in\{15, \ldots 45\}} \Psi_{x, a, T_2} \right) -->
<!-- $$ -->

The observed number of HIV positive and already on ART among ANC clients at $Y[T_1]$ is modelled by
\begin{align*}
Y_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(X_{\{x\} Y[T_1]}^{\text{ANC}}, \rho_{\{x\},\{15, \ldots 45\}, t}^{\text{ANC}}\right) \\
Z_{\{x\} Y[T_1]}^{\text{ANC}} &\sim \text{Bin} \left(Y_{\{x\} Y[T_1]}^{\text{ANC}}, \alpha_{\{x\},\{15, \ldots 45\}, t}^{\text{ANC}}\right)
\end{align*}
where predicted prevalence and ART coverage are aggregated weighted by the predicted number of pregnant women by age $\Psi_{x, a}$
\begin{align*}
\rho_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a}} \\
\alpha_{\{x\}\{a\}, t}^{\text{ANC}} &= \frac{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}} \cdot \alpha_{x, a}^{\text{ANC}}}{\sum_{x \in\{x\}} \sum_{a \in\{a\}} \Psi_{x, a} \cdot \rho_{x, a}^{\text{ANC}}}
\end{align*}
